---
title: "Class 8"
author: "Barry Grant"
date: "10/25/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering

Our first example with **kmeans()** function.

```{r}
# Example plot to see how Rmarkdown works
plot(1:10, typ="l")
```

Back to kmeans...

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```


Use the kmeans() function setting k to 2 and nstart=20 

Inspect/print the results 

```{r}
k <- kmeans(x, centers = 2, nstart = 20)
k
```

Q. How many points are in each cluster?

```{r}
k$size
```

Q. What ‘component’ of your result object details 
      - cluster size?
      - `k$size`
      - cluster assignment/membership? 
      - cluster center?
```{r}
k$cluster
```

```{r}
table(k$cluster)
```

```{r}
k$centers
```


Plot x colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
palette(c("blue", "red"))
plot(x, col=k$cluster)
points(k$centers, col="green", pch=20, cex=3)
```

Q. Repeat for k=3, which has the lower tot.withinss? 

```{r}
k2 <- kmeans(x, centers = 2, nstart = 20)
k3 <- kmeans(x, centers = 3, nstart = 20)
k4 <- kmeans(x, centers = 4, nstart = 20)
k5 <- kmeans(x, centers = 5, nstart = 20)

k2$tot.withinss
k3$tot.withinss
k4$tot.withinss
k5$tot.withinss

plot( c( k2$tot.withinss, k3$tot.withinss, k4$tot.withinss, k5$tot.withinss), typ="b")
```


# Hierarchical clustering in R 

Let's try out the **hclust()** function for Hierarchical clustering in R. This function
need a distance matrix as input!

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```


```{r}
plot(hc)
# Draw a line on the dendrogram
abline(h=8, col="red")

# Cut the tre to yield cluster membership vector
#cutree(hc, h=8)
cutree(hc, k=2)
```


### A more 'real' example of data clustering 

```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2),   # c1
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
  matrix(c(rnorm(50, mean = 1, sd = 0.3),           # c3
           rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")

# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters 
#         (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) ) 

palette( c("red","blue","black"))
plot(x, col=col, pch=20)
```

Or more simple plot by grp
```{r}
grp.names <- rep( c("red", "blue", "black"), each=50)
plot(x, col=grp.names, pch=20)
```


Q. Use the dist(), hclust(), plot() and cutree()
      functions to return 2 and 3 clusters 

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

Use cutree() functions to return 2 and 3 clusters

```{r}
grp.2 <- cutree(hc, k=2)
grp.3 <- cutree(hc, k=3)
plot(x, col=grp.3, pch=20)
```

Q. How does this compare to your known 'col' groups?

```{r}
# Use the table() function!!
table(grp.3, col)
```



# Principal component analysis

Lets make up some data with quite a few dimensions 
```{r}

## Initialize a blank 100 row by 10 column matrix
mydata <- matrix(nrow=100, ncol=10)

## Lets label the rows gene1, gene2 etc. to gene100
rownames(mydata) <- paste("gene", 1:100, sep="")

## Lets label the first 5 columns wt1, wt2, wt3, wt4 and wt5
##   and the last 5 ko1, ko2 etc. to ko5 (for "knock-out") 
colnames(mydata) <- c( paste("wt", 1:5, sep=""),
                       paste("ko", 1:5, sep="") )

## Fill in some fake read counts
for(i in 1:nrow(mydata)) {
  wt.values <- rpois(5, lambda=sample(x=10:1000, size=1))
  ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))

  mydata[i,] <- c(wt.values, ko.values)
}

head(mydata)
```

Note the prcomp() functions wants us to take the transpose of our data so we have fliped rows and cols...

```{r}
head(t(mydata))
```

Now lets try to find "structure" in this data with **prcomp()**

```{r}
pca <- prcomp(t(mydata), scale=TRUE)
pca
```


Make a PC plot of PC1 vs PC2. For this I use the `$x` component of our output (i.e. pca$x).

```{r}
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2])
```

Lets see how well our PCs are doing (i.e. how much varance from the origional data are they capturing)?

```{r}
## Precent variance is often more informative to look at 
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```

Make a "scree-plot" to see the variance captured in each PC

```{r}
barplot(pca.var.per, main="Scree Plot", 
        xlab="Principal Component", ylab="Percent Variation")
```

Lets make our PC plot a bit more useful…

```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```

## PCA of UK food data

The input data file lives here: https://bioboot.github.io/bimm143_F18/class-material/UK_foods.csv

```{r}
inputfile <- "https://bioboot.github.io/bimm143_F18/class-material/UK_foods.csv"
x <- read.csv(inputfile, row.names = 1)
head(x)
```

Summary heatmap
```{r}
heatmap(as.matrix(x))
```

now some PCA
```{r}
pca <- prcomp( t(x) )
summary(pca)
```

Plot our PC results

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))
```





